# Common visualizations in biological analyses

For looking at other common plots, let's take a lot at another data package. This is the [airway](https://bioconductor.org/packages/release/data/experiment/html/airway.html) package, which provides a RangedSummarizedExperiment object of read counts in genes for an RNA-Seq experiment on four human airway smooth muscle cell lines treated with dexamethasone. 

```{r}
library("DESeq2")
library(airway)
library(ggplot2)
library(tidyverse)

data(airway)
se <- airway
dds <- DESeqDataSet(se, design = ~ cell + dex)
rld <- rlog(dds)


keep <- rowSums(counts(dds)) >= 4
dds <- dds[keep,]
dds <- DESeq(dds)
res <- as.data.frame(results(dds))
```
## PCA plot

First we can create a principle component analysis (PCA) of the data. 

```{r}
plotPCA(rld, intgroup = c("dex", "cell"))
```


PCA is a **dimentionality reduction** method. 
It reduced the dimensionality of our data while maintaining as much variation as possible. 
In the above data, it would be impossible to view how each of our samples compare across every gene at the same time. 
PCA finds linear combinations of genes which best explain the variance between each sample.
We can see how much variance is explained by each principle component. 
When examining a PCA plot, we want to make sure that our samples group as expected, mainly, that replicates are closer to eachother than to other samples. 

For other data, we can use the `prcomp` function to perform a PCA analysis. 

```{r}
library(ggfortify)
df <- iris[1:4]
pca_res <- prcomp(df, scale. = TRUE)
autoplot(pca_res, data = iris, colour = 'Species')
```

## Volcano plot

A volcano plot is a common visualization to see the distribution of fold-changes and p values accross our dataset. 
It plots the $log_2$ fold-change against the p values from our statistical analysis. 

```{r}
res %>% 
  mutate(sig = padj < 0.05 & abs(log2FoldChange) > 2) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj), col=sig)) +
       geom_point() +
       geom_vline(xintercept=c(-2, 2), col="red") +
       geom_hline(yintercept=-log10(0.05), col="red")
```

## Heatmap/Clustergram

Heatmaps are a common visualization in a variety of analyses. 
By default in the heatmap package `pheatmap`, the rows and columns of our data are **clustered** using a heirarchical clustering method. 
This allows us to view which samples are most similar, and here how the most differentially expressed genes cluster and change accross different samples. 

```{r}
library(pheatmap)

normalized_counts <- counts(dds, normalized=T) %>% 
                     data.frame() %>%
                     rownames_to_column(var="gene") 
top_genes <- arrange(res, padj) %>%
              slice(1:20)
top_gene_counts <- normalized_counts[match(rownames(top_genes),normalized_counts$gene),]
pheatmap(top_gene_counts[,2:9],
    cluster_rows = T,
    cluster_cols = T,
    show_rownames = T,
    labels_row = top_gene_counts$gene,
    border_color = NA,
    fontsize = 10,
    scale = "row",
    fontsize_row = 10,
    height = 20)
```

## Network visualization

We also sometimes want to visualize networks in R. 
Network visualization can be difficult, and it is reccomended to use programs like [cytoscape](https://cytoscape.org/) for creating publication-ready network figures. 
However, `igraph` is a popular package for handling and visualizing network data in R. 

```{r}
### R code from vignette source 'STRINGdb.Rnw'

library(igraph)

g <- make_ring(10)
plot(g, layout=layout_with_kk, vertex.color="green")
```



------------------------------------------------------------------------

*The materials in this lesson have been adapted from work created by the (HBC)\](http://bioinformatics.sph.harvard.edu/) and Data Carpentry (http://datacarpentry.org/). These are open access materials distributed under the terms of the [Creative Commons Attribution license](https://creativecommons.org/licenses/by/4.0/) (CC BY 4.0), which permits unrestricted use, distribution, and reproduction in any medium, provided the original author and source are credited.*
