# Practice Exercises

::: {.callout-note appearance="simple" icon="false"}
## Basic

In a spreadsheet editor like excel or Google sheets, open the file `../data/messy_temperature_data.csv`.

-   What problems will arise when we load this data into R? If you're unsure, try it out and take a look at the data. Are the columns the types you expected? Does the data appear correct?
-   Inside your spreadsheet editor of choice, fix the problems with the data. Save it under a new file name in your data folder (so that the original data file is not overwritten).
-   Load the dataset into R.
-   What are the dimensions of the dataset? How rows and columns does it have?
:::

::: {.callout-caution appearance="simple" icon="false"}
## Advanced

Try loading the dataset `../data/corrupted_data.txt`. Take a look at the the gene symbols. Some of the gene symbols appear to be dates! [This is actually a common problem in biology.](https://www.nature.com/articles/d41586-021-02211-4)

Try installing the [HGCNhelper](https://cran.r-project.org/web/packages/HGNChelper/index.html) package and using it to correct the date-converted gene symbols.
:::

::: {.callout-important appearance="simple" icon="false"}
## Challenge

As opposed to manually fixing the problems with the dataset from the basic exercise, try to fix the dataset problems using R.
:::

2.  **Working with distributions**

::: {.callout-note appearance="simple" icon="false"}
## Basic

Generate 100 instances of a Poisson(3) random variable.

-   What is the mean?

-   What is the variance as computed by the R function `var`?
:::

::: {.callout-tip icon="false" collapse="true"}
## Solution

```{r}
# Basic
pVars <- rpois(100,3)
mean(pVars)
var(pVars)
```
:::

::: {.callout-caution appearance="simple" icon="false"}
## Advanced

Conduct a binomial test for the following scenario: out of 1 million reads, 19 reads are mapped to a gene of interest, with the probability for mapping a read to that gene being $10^{-5}$.

-   Are these more or less reads than we would expect to be mapped to that gene?

-   Is the finding statistically significant?
:::

::: {.callout-tip icon="false" collapse="true"}
## Solution

```{r}
# Advanced
# Let's check our intuition
table(rbinom(100000, n=1e6, p=1e-6))

# Let's run the test
binom.test(x = 19, n = 1e6, p = 1e-6)
```
:::

::: {.callout-important appearance="simple" icon="false"}
## Challenge

Create a function, `bh_correction`, which takes in a vector of p-values and a target FDR, performs the Benjamini-Hochberg procedure, and returns a vector of p-values which should be rejected at that FDR.
:::

::: {.callout-tip icon="false" collapse="true"}
## Solution

```{r}
# Challenge

bh_correction <- function(pvals, phi){
  pvals <- sort(pvals)
  m <- length(pvals)
  k <- 1
  test_val <- phi/m
  while((test_val>pvals[k]) && (k<m)){
    k <- k+1
    test_val <- (phi*k)/m
  }
  return(pvals[1:k])
}

# Let's test the solution
x <- rnorm(50, mean = c(rep(0, 25), rep(3, 25)))
pvals <- 2*pnorm(sort(-abs(x)))
bh_correction(pvals,0.05)
```
:::
