# Categorical Data in R

## Factors

Since factors are special vectors, the same rules for selecting values using indices apply.

```{r}
expression <- c("high","low","low","medium","high","medium","medium","low","low","low")
```

The elements of this expression factor created previously has following categories or levels: low, medium, and high.

Let's extract the values of the factor with high expression, and let's using nesting here:

```{r}
expression[expression == "high"]    ## This will only return those elements in the factor equal to "high"
```

> **Nesting note**:
>
> The piece of code above was more efficient with nesting; we used a single step instead of two steps as shown below:
>
> Step1 (no nesting): `idx <- expression == "high"`
>
> Step2 (no nesting): `expression[idx]`

## Releveling factors

We have briefly talked about factors, but this data type only becomes more intuitive once you've had a chance to work with it. Let's take a slight detour and learn about how to **relevel categories within a factor**.

To view the integer assignments under the hood you can use `str()`:

```{r}
expression
```

The categories are referred to as "factor levels". As we learned earlier, the levels in the `expression` factor were assigned integers alphabetically, with high=1, low=2, medium=3. However, it makes more sense for us if low=1, medium=2 and high=3, i.e. it makes sense for us to "relevel" the categories in this factor.

To relevel the categories, you can add the `levels` argument to the `factor()` function, and give it a vector with the categories listed in the required order:

```{r}
expression <- factor(expression, levels=c("low", "medium", "high"))     # you can re-factor a factor 
```

Now we have a releveled factor with low as the lowest or first category, medium as the second and high as the third. This is reflected in the way they are listed in the output of `str()`, as well as in the numbering of which category is where in the factor.

> Note: Releveling becomes necessary when you need a specific category in a factor to be the "base" category, i.e. category that is equal to 1. One example would be if you need the "control" to be the "base" in a given RNA-seq experiment.
